Ornaments as indicators of Social change before and after European contact at Kiwulan, Northeastern Taiwan

### 2019-07-09 analysis/paper/paper.Rmd
revise and edit reference

The documents written by Chinese historians in Qing dynasty usually described the decorative purposes of the ornaments and some documents had similar texts that might be copied from previous ones. According to those records, [@Chen2006[1852]: 308], [@Ko1993[1837]: 11, 126] and [@Yao1996[1829]: 77], it is mentioned that indigenous people in Yilan used ornaments in ceremonial contexts to display their status. Amoung those ornaments, golden carp-shaped ornaments made from metal threads had higher value due to its delicacy and rare material that only possessed by rich people. Other people wore agate beads or glass beads on their head or neck to participating various ceremonies. In 1895, at the beginning of Japanese colonization, Ino Kanori [@Kanori1996, pp. 227-232] conducted a field survey for several ethnic groups and found that carp-shaped ornaments had not been used in Yilan that time, but elder people still used beads as ornaments.

Recent ethnographic research of the Kavalan tribe describes that agate beads were used in divination practice, called "subli", by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate bead is valuable object that are usually passed down from mother to daughter as a heirloom in shaman families without knowing the exact origin. Were not been using, beads would be often hided at invisible place such as on the roof beams to make sure no one sees that. During the Japanese colonization period, the number of agate beads declines due to the trade with Japanese who live in Taipei for collecting purpose [@Hsu1992, pp. 22] and carp-shaped ornaments had completely disappeared. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination that is consistent with the observation in European colonial period. It was usually the female shamans conducting divination ritual that passed down from their ancestors. Despite most aborigines today don’t know the origin of agate beads, the value of beads is still be highlighted through oral history and their scarcity. Those historical documents from the early 17th century to the modern ethnography described how local indigenous people use ornaments in local cultural contexts that represents some social roles or status. However, compared to European contact period, there is less mention of beads in Chinese contact period and the description about ornaments is only limited to their dressing culture.

@@

excavate an open area for the first time reaching a surface a bit larger than 109 m2.

Kiwulan \@ref(fig:KWL-map) is located at northern Yilan and was excavated from 2001 to 2004 that creates an open area with 262 squares (4m by 4m) reaching 3,814 meter squares in total. Most beads were found in situ while digging and some were collected through screens with 20 mm and 1.5 mm mesh. The archaeological evidence includes a rich amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures. The chronology of Kiwulan can be divided into Lower Layer Culture (700 - 1200 AD) and Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between based on a series of 32 radiocarbon dates, which coordinate to the late Iron Age and Proto-historical period in Taiwan. This paper focuses on the Upper Layer Culture which overlaps with the contact periods with the Europeans and Chinese.

The earliest record of direct European contact with indigenous people in Yilan can be traced back to 1632, when the local settlements were attacked by the Spanish according to the official documents [add citation; Borao Mateo 2009]. Later in 1647, the Dutch attacked the indigenous villages in Yilan and forced them to accept colonial rules and economic demands by paying annual tribute [add citation; Borao Mateo 2009]. According to Dutch census reports in 1650, Kiwulan was the biggest indigenous community in the Yilan Plain, with a population of 840 adults [@Chen2007; @Kang2012; @Li2014]. The diagnostic artifacts during European contact found at Kiwulan including An-ping jars and glass beads that were largely introduced to Taiwan during the early 17th century.

The European colonization ended in 1662 when they were defeated by the kingdom of Tungning found by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The evidence of Chinese migration can be identified from Chinese official records and large amount of Chinese blue-and-white porcelain found at Kiwulan.

@@

The ornaments studied in this paper came from 40 adjacent 4m x 4m units out of 262 units in the middle part of excavation (Figure \@ref(fig:KWL-sam-area). The sampling principle is based on the extent of intact contexts that the least disturbed by modern constructions are the preference. The Upper Layer Culture can be divided into six sub layers spanning from 14 century to 19 century according to current radiocarbon dates, excavation depth, types of ceramics, and sediment texture and color [@Hsieh2009; @Wang2011]. Layers are assigned numbers from 1 to 6 for which L1 corresponds to the upper context and L6, also the bottom layer, refers to lower context. Based on @Hsieh2009 chronological study of Chinese porcelains, L5 and L6 represent pre-European contact period, L4 was the time of European contact, and L2 and L1 was the period of Chinese contact.

<--add a chart of chronology-->

In general, complete ornaments with clear context were mostly found in burials that helps to understand how people wear them and any relationships between shapes of ornaments and gender or age. In addition to burials, ornaments were also found in middens and living space that can be identified by post holes. The frequency of ornaments from burials is more than those from the living space due to well preserved condition. There is a wide variety of raw materials for ornaments of which ornaments glass, agate and metal ornaments that account for 99% of all ornaments were thought to be introduced by trading with the Europeans and Chinese traders, while ornaments made of shells, woods, and bones were believed locally made that is about only about 1% of all ornaments (Table 1). The reason for the large quantity of glass beads is that they usually have been found in clusters because of the original form such as necklaces. The shape of ornaments include beads, bracelets, rings, bells, pendants, and knitted objects that shows shape are closely related to raw material due to its characteristic. For example, beads are only made of glass, agate, and shell, and bone is the main material for knitted objects that used to connect beads. Also, metal ornaments covers more shapes such bells, pendants, and rings [@Chen2007].

@@

To explore whether body ornaments can reflect the colonial influences on indigenous populations, this research examines the frequency and spatial distribution of trade ornaments at Kiwulan during different time periods with foreign contacts, including European and Chinese contacts. Trade ornaments were analyzed by three analytical units including pre-European contact, European contact, and Chinese contact that indicates three scale of exchange network from less to intense contact subsequently. Because the appearance of ornaments is highly related to the original raw material couple with the function such as bead and bracelet, they are combined as the first criterion of classification which determines five major categories. Under each category of raw material, ornaments are further classified into several subtype according to the shape if any variations were identified in each category. In order to compare the general pattern of changes for major ornaments, less frequent ornaments are combined as one category named other, which will be discussed separately.

Ornaments with clear contexts were mainly found from the post-hole area, burials, and middens, where ornaments from the post-hole area accounts for the majority of the amount in total. Post-hole area is also the contexts that we mainly focus on because it represents house units accross the site that can further provide a basis for comparison to see if there is any spatial pattern of distribution. Ornaments from middens and burials will also be discussed in the later section. For the post-hole area, there were `r sample_size` ornaments recovered from the 40 sampling squares, which accounts for `r sample_pro`% of the total number of ornaments found in the same excavation section including all contexts. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells. Ornaments in most categories can be divided into sub-types according to the variation of shape. The variation was defined by the difference identified in at least two variables, for example dimensional variables such as length, or stylish variables such as patterns. The frequency and feature of subtypes for major categories are summarized in \@ref(tab:tidy-data-for-table).

>>>>

The documents written by Chinese historians in Qing dynasty usually described the decorative purposes of the ornaments and some documents had similar texts that might be copied from previous ones. According to those records, [@Chen2006[1852]: 308], [@Ko1993[1837]: 11, 126] and [@Yao1996[1829]: 77], it is mentioned that indigenous people in Yilan used ornaments in ceremonial contexts to display their status. Amoung those ornaments, golden fish-shaped necklaces made of copper threads had higher value due to its delicacy and more materials invested in production that usually possessed by rich people. Other people wore agate beads or glass beads on their head or neck to participating various ceremonies. In 1895, at the beginning of Japanese colonization, Ino Kanori [@Kanori1996, pp. 227-232] conducted a field survey for several aboriginal groups and found that golden necklaces had not been used in Yilan that time, but elder people still used beads.

Recent ethnographic research of the Kavalan tribe describes that agate beads were used in divination practice, called "subli", by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate bead is valuable object that are usually passed down from mother to daughter as a heirloom in shaman families without knowing the exact origin. Were not been using, beads would be often hided at invisible place such as on the roof beams to make sure no one sees that. During the Japanese colonization period, the number of agate beads declines due to the trade with Japanese who live in Taipei for collecting purpose [@Hsu1992, pp. 22] and golden fish-shaped necklaces had completely disappeared. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination that is consistent with the observation in European colonial period. It was usually the female shamans conducting divination ritual that passed down from their ancestors. Despite most aborigines today don’t know the origin of agate beads, the value of beads is still be highlighted through oral history and their scarcity. Those historical documents from the early 17th century to the modern ethnography described how local indigenous people use ornaments in local cultural contexts that represents some social roles or status. However, compared to European contact period, there is less mention of beads in Chinese contact period and the description about ornaments is only limited to their dressing culture.

@@

Kiwulan \@ref(fig:KWL-map) is located at northern Yilan and was a rescue archaeology projects that carried out from 2001 to 2004 in advance of construction. The total area includes eight open area sections that created 262 squares (4m by 4m) reaching 3,814 meter squares in total [@Chen2007]. Most ornaments were found in situ in the context while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The archaeological evidence includes a rich amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures that indicates it was a settlement. The chronology of Kiwulan can be divided into Lower Layer Culture (700 - 1200 AD) and Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between based on a series of 32 radiocarbon dates [@Chen2007 ]. This paper focuses on the Upper Layer Culture spanning from the late Iron Age that also called proto-historical period [@Liu2017], and historical period starting from the 17th century with European contact in Taiwan.

The earliest record of direct European contact with indigenous people in Yilan can be traced back to 1632, when the local villages were attacked by the Spanish who took revenge on an incident happened earlier that year [@Borao2001, pp. 163]. Later in 1647, the Dutch attacked the indigenous villages and forced them to accept colonial rules and economic demands by paying annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the Yilan Plain, with a population of 840 adults [@Nakamura1938, pp. 12]. The diagnostic artifacts of European contact found at Kiwulan including An-ping jars, stonewares that were largely introduced to Taiwan during the early 17th century. Those artifacts coupled with the original radiocarbon dates were used to identify the pre-European and post-European periods. The European colonization ended in 1662 when they were defeated by the kingdom of Tungning found by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The evidence of Chinese migration can be identified from Chinese official records and large amount of Chinese daily life porcelains found at Kiwulan.

@@

The ornaments studied in this paper found from 40 adjacent square of the largest open area that was located in the middle part of excavation (Figure \@ref(fig:KWL-sam-area). They were sampled because those units were more stratigraphically intact with the least disturbance by modern constructions. The Upper Layer Culture component could be divided into six sub contexts spanning from 14 century to 19 century with a hundred year interval according to original radiocarbon dates, excavation depth, and types of diagnostic porcelain [@Hsieh2009; @Wang2011]. The contexts were assigned numbers from 1 to 6 where L1 corresponds to the youngest context and L6, the bottom layer, refers to the oldest context. To obtain a more reliable chronology for comparison between episodes of culture contact,I reevaluate those contexts for each sampled square by focusing on Anping jars and stonewares that are believed manufactured and distributed in Southeast Asia through European trade route during the early 17th century [@Ketel2011]. Those two categories of ceramic coupled with radiocarbon dates were used to identify the pre-European and post-European periods. Following the previous numbering principle, I separated three episodes of time including pre-European contact period(L5, L6), European contact period(L4), and Chinese contact period (L2 and L1).

In general, complete ornaments with clear context were mostly found in burials that helps to understand how people wear them. In addition to burials, large amount of ornaments were found at household area that were identified by a series of post holes across the space. Here I focus on the ornaments found at post-hole area considering the overall quantity and the comparison across the space. Ornaments were made from a wide variety of raw materials, in which glass, agate and copper are the common materials that were thought to be introduced through trading with the Europeans and Chinese traders since no clear evidence of waste by-products in the archaeological record [@Chen2007]. Although a possibility of small workshops that individuals were able to produce golden beads from prepared glass materials by heating [@Cheng2008], no evidence of accompanying tools or materials in proportion to the large amount of golden beads were found. The shape of ornaments include beads, bracelets, rings, bells, pendants, and knitted objects. It shows that the shape is closely related to what kind of raw material it was made from. For example, beads were only made of glass and agate, while copper ornaments were in the form such as bells, bracelets, and rings [@Chen2007].

@@

To explore whether body ornaments can reflect the colonial influences on indigenous populations, this paper examines the variety, frequency, and spatial distribution of trade ornaments across different time periods, including pre-European contact, European contact, and Chinese contact. Those three analytic units could represent three scales of exchange network from small to large, and three degree of contact from weak to intense. Because the appearance of ornaments is highly related to the original raw material, the type of raw material is the first criterion of classification that determined five major categories. In each category, ornaments were classified into several subtype according to their shapes when variations were present in each category. The variation was defined by the difference identified in at least two variables, for example dimensional variables such as length, or stylish variables such as patterns. In order to compare the general pattern of changes for subtypes, less frequent subtypes were left out from the analysis, but will be discussed separately in later section.

Ornaments from the post-hole area accounts for the majority of the amount in total. Post-hole area provides a clear contexts and basis for comparison to see if there is any spatial pattern of distribution. For the post-hole area, there were `r sample_size` ornaments recovered from the 40 sampling squares, which accounts for `r sample_pro`% of the total number of ornaments found in the same open area section. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells. The variety of subtype and their individual frequency are summarized in \@ref(tab:tidy-data-for-table) in terms of three time episodes. [insert image here]
